---
title: "Steps Count"
author: "Jamal Alikhani"
date: "May 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
directory <- "C:/Users/Jamal/Dropbox/ML-Data-CS materials/Data Science Specialization- Coursera/Reproducible Research/HW-W2/"
setwd(directory)
```

## Loading and preprocessing the data

```{r}
df <- read.csv("activity.csv")
head(df)
str(df)
```

The mean total number of steps taken per day:

```{r}
stepPerDay <- aggregate(steps~date, df, sum)
hist(stepPerDay$steps, xlab = "Steps Per Day", main = "histogram of the total number of steps taken each day")

```

mean and median of the total number of steps taken per day:

```{r}
summary(stepPerDay$steps)

```

## Average daily activity pattern

```{r}
stepsByInterval <- aggregate(steps~interval, df, mean, na.rm=TRUE)
with(stepsByInterval, plot(interval, steps , type="l", xlab = "interval (min)", ylab = "Average step (count)", main = "average daily activity pattern"))
```

As per above graph, the maximum number of steps happens in `r with(stepsByInterval, interval[which.max(steps)])` min. 

## Imputing missing values

The total number of missing values are:
```{r}
totalNa <- sum(is.na(df$steps))
```
`r totalNa`.   

In order to check the bias toward removing the missing rows, the daily average value is imputed.

```{r}
dfFilled <- transform(df, steps = ifelse(is.na(df$steps), stepsByInterval$steps[match(df$interval, stepsByInterval$interval)], df$steps)) 
head(dfFilled)
```

To see the effect of missing vaue imputation, the total number of steps taken each day is compare with the data when missing value are just removed. 

```{r}
stepsPerDayImputed <- aggregate(steps~date, dfFilled, sum)

hist(stepsPerDayImputed$steps, col = "blue", xlab = "Steps Per Day", main = "Histogram of the total number of steps taken each day")
hist(stepPerDay$steps, col = "green", add = TRUE, xlab = "Steps Per Day")
legend("topleft", c("Removed NAs", "Imputed NAs"), col = c("green", "blue"), lwd = 10)

```
mean and median of the total number of steps taken per day for the imputed data:

```{r}
summary(stepsPerDayImputed$steps)

```


## differences in activity patterns between weekdays and weekends

Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day:

```{r}
dfFilled$weekDay <- as.factor(ifelse(is.element(weekdays(as.Date(dfFilled$date)),c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")), "weekday", "weekend"))

stepsByInterval <- aggregate(steps ~ interval + weekDay, dfFilled, mean)

library(lattice)
xyplot(stepsByInterval$steps ~ stepsByInterval$interval|stepsByInterval$weekDay,xlab = "interval (min)", ylab = "Average step (count)", main = "average daily activity pattern for Weekday vs. Weekend", layout=c(1,2), type="l")

```
